Limited predictability of amino acid substitutions in seasonal influenza viruses

### 2020-07-01 article.tex
Add A/H1N1 and NA

\newcommand{\RAN}[1]{{\color{blue}Richard: #1}}
\newcommand{\pierre}[1]{{\color{red}Pierre: #1}}
\newcommand{\sref}[1]{S\ref{#1}}

@@

	In this work, we use A/H3N2 HA and NA sequences from year 2000 to 2019 to perform a retrospective analysis of frequency trajectories of amino acid mutations. 
	We quantify how rapidly mutations at different frequencies are lost or fixed and how rapidly they spread through the population.
	We further investigate whether any properties or statistics are predictive of whether a particular mutation fixes or not. 
	To our surprise, we find that the predictability of these trajectories is very limited: The probability that a mutation fixes differs little from its current frequency as would be expected if fixation happened purely by chance.
	This observation holds for many different categories of mutations, including mutations at epitope positions. 
	This quasi-neutrality is not attributable solely to clonal interference and genetic linkage, as simulation of models including even strong interference retain clear signatures of predictability. 
	Consistent with these observations, we show that a simple predictor uninformed by fitness, the consensus sequence, performs as the well as the Local Branching Index (LBI), the growth measure based on the genealogy used in \cite{neher_predicting_2014}. 
	This suggests that although LBI has predictive power, the reason for its success may not be related to it approximating fitness of strains.

@@

The main underlying question asked in this work is the following: given a mutation $X$ in the genome of A/H3N2 influenza that we observe at a frequency $f$ in the population at a given date, what can we say about the future of $X$?  \\
There are different ways to make this question more specific. First, one can try to quantitatively predict the frequency of $X$ at future times $f(t)$. In other words, having observed a mutation at frequencies $(f_1, f_2,\ldots,f_n)$ at dates $(t_1,t_2,\ldots,t_n)$, what can we say about its frequency at future dates $(t_{n+1}, t_{n+2},\ldots)$?
A simpler, more qualitative question, is to ask whether $X$ will fix in the population, will disappear, or whether the site will stay polymorphic.

In this work, we use amino-acid sequences of the HA and NA genes of A/H3N2 influenza since the year 2000 available in GISAID \citep{shu2017gisaid} (see supplementary materials for an acknowledgment of all data contributors). 
This amounts to $44\,976$ HA and $36\,300$ NA sequences, with a minimum of 100 per year. These sequences are binned in non-overlapping intervals of one month. Each single-month time bin and the sequences that it contains represent a (noisy) snapshot of the A/H3N2 population at a given date. The number of sequences per time bin varies strongly both with year and according to the season, with earlier time bins containing around 10 sequences while more recent bins contain several hundreds (see figures~\sref{fig:nseq_per_year} and \sref{fig:nseq_per_month} in SM for details).

@@

To answer this question retrospectively, we use all frequency trajectories extracted from A/H3N2 HA sequences that satisfy these conditions for a given $f_0$.  The number of trajectories is limited and the frequency estimates themselves are based on a finite sample and are hence imprecise. 
Therefore, we consider trajectories in an interval $[f_0-\delta f, f_0+\delta f]$ with $\delta f = 0.05$.

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel2.png}
	\caption{\textbf{A}: Sketch of the idea behind the short term prediction of frequency trajectories. Given a mutation that we have seen increasing in frequency and that we ``catch'' at frequency $f_0$ at time $t_0$, what can we say about the distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$? \textbf{B}: All frequency trajectories of amino acid mutations in the HA gene that were absent in the past, are seen around $f_0=30\%$ frequency at time $t_0=0$, and are based on more than 10 sequences at each time point. Red curves represent mutations that will ultimately fix, blue the ones that will be lost, and black the ones for which we do not know the final status. Dashed horizontal lines (blue and red) represent loss and fixation thresholds. The thick black line is the average of all trajectories, counting those that fix (resp. disappear) as being at frequency 1 (resp. 0). \textbf{C}: Distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$ for the trajectories shown in panel \textbf{B} and for specific values of $\Delta t$. }
	\label{fig:trajectories}
\end{figure*}

\subsection*{Prediction of fixation or loss}
Instead of predicting future frequency, let's consider the more modest goal of predicting the probability that a mutation fixes in the population. 
We first estimate the fraction of frequency trajectories that either fix in the population or are lost, as well as the time it takes for one or the other to happen. 
Panel \textbf{A} of figure~\ref{fig:fix_loss_active} shows the fraction of frequency trajectories in HA that either have fixed, were lost or remained active as a function of the time elapsed since they were first seen above 25\% frequency. 
Most mutations are either lost or become fixed after 2-3 years, with very few trajectories remaining active after 5 years.
This time scale of 2-3 years is consistent with the typical coalescence time observed in phylogenetic trees of A/H3N2 influenza \citep{rambaut_genomic_2008,yan_phylodynamic_2019}. 
We also note that the fraction of lost trajectories increases sharply at small times, indicating that 40\% of mutations observed above 25\% frequency are lost within one year, while it takes longer to fix a mutation in the whole population.

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel3.png}
	\caption{\textbf{A}: Activity of all rising frequency trajectories seen above 25\% frequency. \textbf{B}: Probability of fixation of a mutation (amino acid or synonymous) $P_{fix}(f)$ as a function of the frequency $f$ at which it is measured. Only new mutations are considered, \emph{i.e.} mutations that were absent in the past. The diagonal dashed line is the expectation from a neutrally evolving population. Colored dashed lines represent synonymous mutations. Colored solid lines represent amino acid mutations. Error bars represent a 95\% confidence interval.}
	\label{fig:fix_loss_active}
\end{figure*}

We then examined the probability of mutations to fix in the population as a function of the frequency at which they are seen. 
For different values of frequency $f$, we consider all trajectories that started at a null frequency and are seen in the interval $[f - 7.5\%, f + 7.5\%]$ at any given time. 
The probability of a mutation fixing given that it is seen at frequency $f$, $P_{fix}(f)$, is then estimated by the fraction of those trajectories which terminate at a frequency larger than $95\%$, \emph{i.e.} our fixation threshold.
Panel \textbf{B} of figure~\ref{fig:fix_loss_active} shows $P_{fix}(f)$ as a function of $f$ for NA and HA. 
For both these proteins, the probability of fixation of a new mutation at frequency $f$ is very close to $f$ itself, that is $P_{fix}(f)\simeq f$. 
This result is exactly what is expected in a population evolving in the \emph{absence} of selection. 
Indeed, in a neutrally evolving and structure-less population of constant size $N$ such as in the Wright-Fisher model~\cite{kimura_diffusion_1964}, the probability that any individual ultimately becomes the ancestor of all the future population is $1/N$. 
A mutation or trait appearing at frequency $f$ is shared by $f\cdot N$ individuals, and the probability for one of them to become the ancestor of all the future population is $f\cdot N/N=f$. 
Thus, the probability of this mutation or trait to fix in the population is equal to its current frequency, a case which we will refer to as the neutral expectation. 
Panel \textbf{B} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation.

This quasi-neutral dynamics is in apparent contradiction with evidence that influenza surface proteins are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
If strong selection was present, we would expect rising amino acid mutations to fix at a higher frequency than the one at which they are measured. 
In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.

To investigate further, we try to find categories of mutations that deviate from the neutral expectation of figure~\ref{fig:fix_loss_active}. 
We first turn to the \emph{Local Branching Index} (LBI), a quantity calculated for each node in a phylogenetic tree that indicates how dense the branching of the tree is around that node. 
LBI has previously been successfully used as a predictor of the future population of influenza~\cite{neher_predicting_2014}, and was shown to be a proxy for fitness of leaves or ancestral nodes in mathematical models of evolution. 
Here, we define the LBI of a mutation at date $t$ as the average LBI of strains that carry this mutation and that were sampled in the time bin corresponding to $t$. 
Panel A of figure~\ref{fig:pfix_stratified} shows fixation probability for HA mutations with LBI in the top or bottom half of the distribution. 
Both groups are again consistent with neutral evolution, suggesting that LBI carries very little information on the probability of fixation of a mutation.

Next, we focused on previously reported antigenic sites in the HA protein, referred to as \emph{epitope} positions. 
Mutations at these position might mediate immune escape and are therefore likely under strong selection. 
We used four lists of relevant epitope positions from different sources comprising from 7 to 129 positions in the sequence of the HA1 protein~\cite{Koel976, luksza_predictive_2014, Shih6283, wolf_long_2006}. 
Panel Fig.~\ref{fig:pfix_stratified}\textbf{B} shows fixation probability as a function of frequency for the four lists of epitopes. 
Only mutations at the 7 epitope sites reported in~\cite{Koel976} have higher chances of fixation than expected by chance. 
No clear difference is found for the lists by \citet{luksza_predictive_2014, wolf_long_2006}, while positions from \citet{Shih6283} show lower chances of fixation.
%However, even in the former case of the positions from \citet{Koel976}, observations are different from what one would expect if the dynamics where dominated by selective sweeps. 
One should also note that many of these positions were determined post-hoc and enriched for positions that experienced rapid substitutions before the publication of the respective studies.

@@

Since influenza is seasonal in temperate regions, geographic spread and persistence might be predictive of the success of mutations. 
We quantify geographic spread of a mutation by the entropy of its frequency distribution across regions (see methods) and its persistence by the age of the trajectory by the time it reaches frequency $f$. 
Figure~\sref{fig:geospread_and_time} shows the fixation probabilities as a function of observed frequency for mutations classified according to these scores. 
The two scores also allow a quantitatively moderate distinction between mutations: for a given frequency $f$, mutations found in many regions or those that are older (in the sense that they have taken more time to reach frequency $f$) tend to fix more often than geographically localized mutations or more recent ones, but the effect is small.
These two scores are in fact correlated, with older trajectories representing mutations that are more geographically spread, as can be seen in figure~\sref{fig:geospread_v_time} of SM.
% A natural way to explain these observations would be to refer to the seasonal characteristics of A/H3N2. 
% In temperate regions, the number of A/H3N2 infections drastically increases during winter, leading to an at first rapidly growing and then declining population of viruses. 
% For this reason, one could imagine that some mutations rise in frequency in an ``opportunistic'' way during an epidemic season, but do not survive the following decline in population size. 
% For the same reason, mutations that are localized in a particular region might mostly be due to the epidemic season specific to that region. 
However, it is important to note that sampling biases and heterogeneity across time and space (see supplementary figures~\sref{fig:nseq_per_year} and \sref{fig:nseq_per_month}) make answering such specific hypothesis challenging.
Frequency of mutations might thus be amplified through different sampling biases, making the connection between geographic spread, seasonality and mutation frequency non-trivial to measure. 
%It is however interesting to observe that quantities based on the population's geographical and temporal structure can apparently give as much or more information on fixation probability and predictability than genomic or phylogenetic ones.

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel4.png}
	\caption{Fixation probability $P_{fix}(f)$ as a function of frequency. \textbf{A}: Mutation with higher or lower LBI values, based on their position with respect to the median LBI value. \textbf{B}: Different lists of epitope positions in the HA protein. The authors and the number of positions is indicated in the legend. \textbf{C}: Mutations for binary positions, \emph{i.e.} positions for which we never see more than two amino acids in the same time bin. \textbf{D}: Mutations that appear once or more than once in the tree for a given time bin.}
	\label{fig:pfix_stratified}
\end{figure*}

@@

The results shown in figures~\ref{fig:fix_loss_active} and \ref{fig:pfix_stratified} are difficult to reconcile with the notion that seasonal influenza virus evolution is driven by rapid directed positive selection. 
One possible explanation for the quasi-neutral behaviour of mutations might be tight genetic linkage inside each protein and strong competition between different adaptive mutations \citep{neher_genetic_2011}. 
We design a simple model of population evolution based on the \texttt{ffpopsim} simulation software to test this hypothesis~\cite{10.1093/bioinformatics/bts633}. 
The model represents a population of binary genomes of length $L=200$ evolving in a fitness landscape that changes through time.

@@

These models are not meant to be accurate models of influenza viruses evolution. 
But figure~\sref{fig:simulations} does show is that the patterns observed in influenza virus evolution are not readily reproduced by in models of adapting populations, even if they include strong clonal competition and epistasis.
We conclude that the pattern in figure~\ref{fig:fix_loss_active} is not a straightforward manifestation of genetic linkage and clonal interference, but that some more intricate interplay of epidemiology, seasonality, human immunity and chance gives rise to the quasi-neutral yet strongly selected evolutionary dynamics of IAVs.

@@

To quantify the ability of the LBI and other measures to pick good representatives of the future, we construct a large tree of HA sequences with 100 sequences in non-overlapping time bins of 4 months from year 2003 to 2019 (a total of 4402 as some 4 month intervals contain less than 100 sequences).
Each time bin is considered as a snapshot of the A/H3N2 influenza population and we will refer to sequences in time bin $t$ as the population of the \emph{present}. 
From this present population, we will predict \emph{future} populations in time bin $t+\Delta t$, using only sequences in time bin $t$ and before.

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel5.png}
	\caption{\textbf{A}: Average Hamming distance of the sequences of different predictors to HA sequences of future influenza populations, themselves averaged over all ``present'' populations from year 2003 to 2019. Predictors are: a randomly picked sequence in the present population; the sequence of the strain with the highest LBI in the present population; the consensus sequence of the present population. \textbf{B}: Scaled Hamming distance between the sequence of the top LBI strain and the consensus sequence for populations at different dates. The scaling is such that for each date, the Hamming distance between a strain from the population and the consensus is on average $1$. The strain with the highest LBI is almost always closer to the consensus sequence than the average strain.}
	\label{fig:LBI_consensus}
\end{figure*}

@@

The premises of predicting the trajectory of a mutation are (i) significant fitness difference between genomes carrying different variants at the site and (ii) a selection pressure that changes slowly over time. 
Under such conditions, it is expected that frequency trajectories will show an persistent behavior which would make them predictable for some time. 
However, we could find only limited evidence for such persistent behavior in the past 19 years of A/H3N2 evolution.
This lead us to conclude that (i) influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) previous methods to predict influenza evolution work primarily because they pick strains that represent the future well, not because they predict future dynamics.

The primary quantity we investigated is the probability that a novel mutation fixes, given it is currently observed at frequency $f$.
In neutral models of evolution this probability is equal to $f$, while it should be higher or lower than $f$ for beneficial or deleterious mutations, respectively \citep{kimura_diffusion_1964}.
This effect is readily seen in simulations of simple models of evolution, but we failed to identify groups of substitutions in which fixation probability deviates substantially from their instantaneous frequency. 
In figure \ref{fig:pfix_stratified}, we tried to break trajectories into groups that where previously associated with rapid adaptation (e.g.~epitope positions as defined in \citep{Shih6283,Koel976,bush_predicting_1999,luksza_predictive_2014}) or high and low fitness \citep{neher_predicting_2014}. 
Only mutations at the seven ``Koel''-positions fixed more often than expected by chance.
The only other stratification that had a measurable effect on fixation were was whether or not convergent evolution (at the level of amino acid or position) was observed. 
On average, the direction of a frequency trajectory does not persist for longer than a few months and memory of the initial frequency is lost after one year.

This failure to meaningfully predict mutation frequency trajectories is seemingly at odds with the strong signatures of selection in influenza surface proteins \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
These signatures clearly indicate that surface proteins change their amino acid sequence to evade human immunity and these pressures can be recapitulated both in vivo and in vitro \citep{lee_mapping_2019,xue_within-host_nodate}.
A recent preprint proposed that influenza virus evolution is primarily limited by an asynchrony between population level selection and generation of new variants within infected hosts \citep{morris_asynchrony_2020}. 
Along these lines, it is possible that the A/H3N2 population readily responds once population level selection is high enough by giving rise to essentially equivalent variants.
In addition, every successful variant will produce population immunity limiting its future spread. 
These considerations might explain the disconnect between models of rapid adaptation and the frequency dynamics observed in influenza virus populations.

@@

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./SM_figures/cutoff_159S_p2.png}
	\caption{Equivalent to figure \ref{fig:pfix_stratified} of the main text, but with strains carrying the HA1 159S mutation removed.}
	\label{fig:cutoff_159S_p2}
\end{figure*}

\newpage
\section{H1N1 influenza}

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./SM_figures/panel2_h1n1pdm.png}
	\caption{Equivalent of figure \ref{fig:fix_loss_active} of the main text for H1N1 influenza. \textbf{A}: Activity of all rising frequency trajectories seen above 25\% frequency. \textbf{B}: Probability of fixation of a mutation (amino acid or synonymous) $P_{fix}(f)$ as a function of the frequency $f$ at which it is measured. Only new mutations are considered, \emph{i.e.} mutations that were absent in the past. The diagonal dashed line is the expectation from a neutrally evolving population. Colored dashed lines represent synonymous mutations. Colored solid lines represent amino acid mutations. Error bars represent a 95\% confidence interval.}
	\label{fig:fix_loss_active_h1n1}
\end{figure*}

>>>>

\newcommand{\RAN}[1]{{\color{blue}Richard: #1}}
\newcommand{\pierre}[1]{{\color{red}Pierre: #1}}
\newcommand{\discuss}[1]{{\color{green}Todiscuss: #1}}
\newcommand{\sref}[1]{S\ref{#1}}

@@

	In this work, we use A/A/H3N2 HA and NA sequences from year 2000 to 2019 to perform a retrospective analysis of frequency trajectories of amino acid mutations. 
	We quantify how rapidly mutations at different frequencies are lost or fixed and how rapidly they spread through the population.
	We further investigate whether any properties or statistics are predictive of whether a particular mutation fixes or not. 
	To our surprise, we find that the predictability of these trajectories is very limited: The probability that a mutation fixes differs little from its current frequency as would be expected if fixation happened purely by chance.
	This observation holds for many different categories of mutations, including mutations at epitope positions. 
	This quasi-neutrality is not attributable solely to clonal interference and genetic linkage, as simulation of models including even strong interference retain clear signatures of predictability. 
	Consistent with these observations, we show that a simple predictor uninformed by fitness, the consensus sequence, performs as the well as the Local Branching Index (LBI), the growth measure based on the genealogy used in \cite{neher_predicting_2014}. 
	This suggests that although LBI has predictive power, the reason for its success may not be related to it approximating fitness of strains.

@@

The main underlying question asked in this work is the following: given a mutation $X$ in the genome of A/A/H3N2 influenza that we observe at a frequency $f$ in the population at a given date, what can we say about the future of $X$?  \\
There are different ways to make this question more specific. First, one can try to quantitatively predict the frequency of $X$ at future times $f(t)$. In other words, having observed a mutation at frequencies $(f_1, f_2,\ldots,f_n)$ at dates $(t_1,t_2,\ldots,t_n)$, what can we say about its frequency at future dates $(t_{n+1}, t_{n+2},\ldots)$?
A simpler, more qualitative question, is to ask whether $X$ will fix in the population, will disappear, or whether the site will stay polymorphic.

In this work, we use amino-acid sequences of the HA and NA genes of A/A/H3N2 influenza since the year 2000 available in GISAID \citep{shu2017gisaid} (see supplementary materials for an acknowledgment of all data contributors). 
This amounts to $44\,976$ HA and $36\,300$ NA sequences, with a minimum of 100 per year. These sequences are binned in non-overlapping intervals of one month. Each single-month time bin and the sequences that it contains represent a (noisy) snapshot of the A/A/H3N2 population at a given date. The number of sequences per time bin varies strongly both with year and according to the season, with earlier time bins containing around 10 sequences while more recent bins contain several hundreds (see figures~\sref{fig:nseq_per_year} and \sref{fig:nseq_per_month} in SM for details).

@@

To answer this question retrospectively, we use all frequency trajectories extracted from A/A/H3N2 HA sequences that satisfy these conditions for a given $f_0$.  The number of trajectories is limited and the frequency estimates themselves are based on a finite sample and are hence imprecise. 
Therefore, we consider trajectories in an interval $[f_0-\delta f, f_0+\delta f]$ with $\delta f = 0.05$.

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel1.png}
	\caption{\textbf{A}: Sketch of the idea behind the short term prediction of frequency trajectories. Given a mutation that we have seen increasing in frequency and that we ``catch'' at frequency $f_0$ at time $t_0$, what can we say about the distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$? \textbf{B}: All frequency trajectories of amino acid mutations in the A/H3N2 HA and NA genes that were absent in the past, are seen around $f_0=30\%$ frequency at time $t_0=0$, and are based on more than 10 sequences at each time point. Red curves represent mutations that will ultimately fix, blue the ones that will be lost, and black the ones for which we do not know the final status. Dashed horizontal lines (blue and red) represent loss and fixation thresholds. The thick black line is the average of all trajectories, counting those that fix (resp. disappear) as being at frequency 1 (resp. 0). \textbf{C}: Distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$ for the trajectories shown in panel \textbf{B} and for specific values of $\Delta t$. }
	\label{fig:trajectories}
\end{figure*}

\subsection*{Prediction of fixation or loss}
Instead of predicting future frequency, let's consider the more modest goal of predicting the probability that a mutation fixes in the population. 
We first estimate the fraction of frequency trajectories that either fix in the population or are lost, as well as the time it takes for one or the other to happen. 
Panels \textbf{A} and \textbf{B} of figure~\ref{fig:fix_loss_active} shows the fraction of frequency trajectories in HA and NA that either have fixed, were lost or remained active as a function of the time elapsed since they were first seen above 25\% frequency. 
Most mutations are either lost or become fixed after 2-3 years, with very few trajectories remaining active after 5 years.
This time scale of 2-3 years is consistent with the typical coalescence time observed in phylogenetic trees of A/A/H3N2 influenza \citep{rambaut_genomic_2008,yan_phylodynamic_2019}. 
We also note that the fraction of lost trajectories increases sharply at small times, indicating that 40\% of mutations observed above 25\% frequency are lost within one year, while it takes longer to fix a mutation in the whole population.

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel2.png}
	\caption{\textbf{A}: Activity of all rising frequency trajectories seen above 25\% frequency for A/H3N2 HA and NA. \textbf{B}: Same as \textbf{A} for A/H1N1. \textbf{C}: Probability of fixation of a mutation (amino acid or synonymous) $P_{fix}(f)$ as a function of the frequency $f$ at which it is measured, for A/H3N2 HA and NA. Only new mutations are considered, \emph{i.e.} mutations that were absent in the past. The diagonal dashed line is the expectation from a neutrally evolving population. Colored dashed lines represent synonymous mutations. Colored solid lines represent amino acid mutations. Error bars represent a 95\% confidence interval. \textbf{D}: Same as \textbf{C} for A/H1N1.}
	\label{fig:fix_loss_active}
\end{figure*}

We then examined the probability of mutations to fix in the population as a function of the frequency at which they are seen. 
For different values of frequency $f$, we consider all trajectories that started at a null frequency and are seen in the interval $[f - 7.5\%, f + 7.5\%]$ at any given time. 
The probability of a mutation fixing given that it is seen at frequency $f$, $P_{fix}(f)$, is then estimated by the fraction of those trajectories which terminate at a frequency larger than $95\%$, \emph{i.e.} our fixation threshold.
Panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active} shows $P_{fix}(f)$ as a function of $f$ for NA and HA and for influenza A/H3N2 and A/H1N1. 
\discuss{For both these proteins, the probability of fixation of a new mutation at frequency $f$ is very close to $f$ itself, that is $P_{fix}(f)\simeq f$.} 
This result is exactly what is expected in a population evolving in the \emph{absence} of selection. 
Indeed, in a neutrally evolving and structure-less population of constant size $N$ such as in the Wright-Fisher model~\cite{kimura_diffusion_1964}, the probability that any individual ultimately becomes the ancestor of all the future population is $1/N$. 
A mutation or trait appearing at frequency $f$ is shared by $f\cdot N$ individuals, and the probability for one of them to become the ancestor of all the future population is $f\cdot N/N=f$. 
Thus, the probability of this mutation or trait to fix in the population is equal to its current frequency, a case which we will refer to as the neutral expectation. 
Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/A/H3N2 influenza are in good agreement with the neutral expectation. \discuss{Add smth for A/H1N1}

\discuss{This quasi-neutral dynamics is in apparent contradiction with evidence that influenza surface proteins are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
If strong selection was present, we would expect rising amino acid mutations to fix at a higher frequency than the one at which they are measured. 
In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.}

To investigate further, we try to find categories of mutations that deviate from the neutral expectation of panel \textbf{C} of figure~\ref{fig:fix_loss_active}. 
We first turn to the \emph{Local Branching Index} (LBI), a quantity calculated for each node in a phylogenetic tree that indicates how dense the branching of the tree is around that node. 
LBI has previously been successfully used as a predictor of the future population of influenza~\cite{neher_predicting_2014}, and was shown to be a proxy for fitness of leaves or ancestral nodes in mathematical models of evolution. 
Here, we define the LBI of a mutation at date $t$ as the average LBI of strains that carry this mutation and that were sampled in the time bin corresponding to $t$. 
Panel A of figure~\ref{fig:pfix_stratified} shows fixation probability for HA mutations with LBI in the top or bottom half of the distribution. 
Both groups are again consistent with neutral evolution, suggesting that LBI carries very little information on the probability of fixation of a mutation.

Next, we focused on previously reported antigenic sites in the HA protein, referred to as \emph{epitope} positions. 
Mutations at these position might mediate immune escape and are therefore likely under strong selection. 
We used four lists of relevant epitope positions from different sources comprising from 7 to 129 positions in the sequence of the HA1 protein~\cite{Koel976, luksza_predictive_2014, Shih6283, wolf_long_2006}. 
Panel Fig.~\ref{fig:pfix_stratified}\textbf{B} shows fixation probability as a function of frequency for the four lists of epitopes. 
Only mutations at the 7 epitope sites reported in~\cite{Koel976} have higher chances of fixation than expected by chance. 
No clear difference is found for the lists by \citet{luksza_predictive_2014, wolf_long_2006}, while positions from \citet{Shih6283} show lower chances of fixation.
One should also note that many of these positions were determined post-hoc and enriched for positions that experienced rapid substitutions before the publication of the respective studies.

@@

\discuss{We conduct the same analysis on A/H1N1 influenza, with results visible in figure \sref{fig:pfix_stratified_h1n1}. 
Results are qualitatively similar to those obtained for A/H3N2, with LBI giving little information and mutations at non-binary positions having a higher chance of fixation. 
Panel \textbf{D} differs between figures~\ref{fig:pfix_stratified} and \sref{fig:pfix_stratified_h1n1}, with convergent evolution giving less information on fixation in the latter case. 
However, this could be due to the shorter time period over which A/H1N1 evolved, resulting in a shorter tree and less possibilities of convergent evolution. 
Indeed, error bars for mutations appearing multiple times in \textbf{D} of figure~\sref{fig:pfix_stratified_h1n1} are relatively large, indicating a lower amount of trajectories.}

Since influenza is seasonal in temperate regions, geographic spread and persistence might be predictive of the success of mutations. 
We quantify geographic spread of a mutation by the entropy of its frequency distribution across regions (see methods) and its persistence by the age of the trajectory by the time it reaches frequency $f$. 
Figures~\sref{fig:geospread_and_time_h3n2} and \sref{fig:geospread_and_time_h1n1pdm} show the fixation probabilities as a function of observed frequency for mutations classified according to these scores. 
The two scores also allow a quantitatively moderate distinction between mutations: for a given frequency $f$, mutations found in many regions or those that are older (in the sense that they have taken more time to reach frequency $f$) tend to fix more often than geographically localized mutations or more recent ones, but the effect is small.
These two scores are in fact correlated, with older trajectories representing mutations that are more geographically spread, as can be seen in figure~\sref{fig:geospread_v_time} of SM.
% A natural way to explain these observations would be to refer to the seasonal characteristics of A/A/H3N2. 
% In temperate regions, the number of A/A/H3N2 infections drastically increases during winter, leading to an at first rapidly growing and then declining population of viruses. 
% For this reason, one could imagine that some mutations rise in frequency in an ``opportunistic'' way during an epidemic season, but do not survive the following decline in population size. 
% For the same reason, mutations that are localized in a particular region might mostly be due to the epidemic season specific to that region. 
However, it is important to note that sampling biases and heterogeneity across time and space (see supplementary figures~\sref{fig:nseq_per_year} and \sref{fig:nseq_per_month}) make answering such specific hypothesis challenging.
Frequency of mutations might thus be amplified through different sampling biases, making the connection between geographic spread, seasonality and mutation frequency non-trivial to measure. 
%It is however interesting to observe that quantities based on the population's geographical and temporal structure can apparently give as much or more information on fixation probability and predictability than genomic or phylogenetic ones.

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel3.png}
	\caption{Fixation probability $P_{fix}(f)$ as a function of frequency. \textbf{A}: Mutation with higher or lower LBI values, based on their position with respect to the median LBI value. \textbf{B}: Different lists of epitope positions in the HA protein. The authors and the number of positions is indicated in the legend. \textbf{C}: Mutations for binary positions, \emph{i.e.} positions for which we never see more than two amino acids in the same time bin. \textbf{D}: Mutations that appear once or more than once in the tree for a given time bin.}
	\label{fig:pfix_stratified}
\end{figure*}

@@

\discuss{The results shown in figures~\ref{fig:fix_loss_active} and \ref{fig:pfix_stratified} are difficult to reconcile with the notion that seasonal influenza virus evolution is driven by rapid directed positive selection.} 
One possible explanation for the quasi-neutral behaviour of mutations might be tight genetic linkage inside each protein and strong competition between different adaptive mutations \citep{neher_genetic_2011}. 
We design a simple model of population evolution based on the \texttt{ffpopsim} simulation software to test this hypothesis~\cite{10.1093/bioinformatics/bts633}. 
The model represents a population of binary genomes of length $L=200$ evolving in a fitness landscape that changes through time.

@@

These models are not meant to be accurate models of influenza viruses evolution. 
\discuss{But figure~\sref{fig:simulations} does show is that the patterns observed in influenza virus evolution are not readily reproduced by in models of adapting populations, even if they include strong clonal competition and epistasis.
We conclude that the pattern in figure~\ref{fig:fix_loss_active} is not a straightforward manifestation of genetic linkage and clonal interference, but that some more intricate interplay of epidemiology, seasonality, human immunity and chance gives rise to the quasi-neutral yet strongly selected evolutionary dynamics of IAVs.}

@@

To quantify the ability of the LBI and other measures to pick good representatives of the future, we construct a large tree of HA sequences with 100 sequences in non-overlapping time bins of 4 months from year 2003 to 2019 (a total of 4402 as some 4 month intervals contain less than 100 sequences).
Each time bin is considered as a snapshot of the A/A/H3N2 influenza population and we will refer to sequences in time bin $t$ as the population of the \emph{present}. 
From this present population, we will predict \emph{future} populations in time bin $t+\Delta t$, using only sequences in time bin $t$ and before.

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel4.png}
	\caption{\textbf{A}: Average Hamming distance of the sequences of different predictors to HA sequences of future influenza populations, themselves averaged over all ``present'' populations from year 2003 to 2019. Predictors are: a randomly picked sequence in the present population; the sequence of the strain with the highest LBI in the present population; the consensus sequence of the present population. \textbf{B}: Scaled Hamming distance between the sequence of the top LBI strain and the consensus sequence for populations at different dates. The scaling is such that for each date, the Hamming distance between a strain from the population and the consensus is on average $1$. The strain with the highest LBI is almost always closer to the consensus sequence than the average strain.}
	\label{fig:LBI_consensus}
\end{figure*}

@@

The premises of predicting the trajectory of a mutation are (i) significant fitness difference between genomes carrying different variants at the site and (ii) a selection pressure that changes slowly over time. 
Under such conditions, it is expected that frequency trajectories will show a persistent behavior which would make them predictable for some time. 
However, we could find only limited evidence for such persistent behavior in the past 19 years of A/A/H3N2 evolution. \pierre{But we do find evidence for A/H1N1, not of a large magnitude.}
This lead us to conclude that (i) A/H3N2 influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) previous methods to predict influenza evolution work primarily because they pick strains that represent the future well, not because they predict future dynamics.

The primary quantity we investigated is the probability that a novel mutation fixes, given it is currently observed at frequency $f$.
In neutral models of evolution this probability is equal to $f$, while it should be higher or lower than $f$ for beneficial or deleterious mutations, respectively \citep{kimura_diffusion_1964}.
\discuss{This effect is readily seen in simulations of simple models of evolution, but we failed to identify groups of substitutions in which fixation probability deviates substantially from their instantaneous frequency.} 
In figure \ref{fig:pfix_stratified}, we tried to break trajectories into groups that where previously associated with rapid adaptation (e.g.~epitope positions as defined in \citep{Shih6283,Koel976,bush_predicting_1999,luksza_predictive_2014}) or high and low fitness \citep{neher_predicting_2014}. 
Only mutations at the seven ``Koel''-positions fixed more often than expected by chance.
The only other stratification that had a measurable effect on fixation were was whether or not convergent evolution (at the level of amino acid or position) was observed. 
On average, the direction of a frequency trajectory does not persist for longer than a few months and memory of the initial frequency is lost after one year.

This failure to meaningfully predict mutation frequency trajectories is seemingly at odds with the strong signatures of selection in influenza surface proteins \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
These signatures clearly indicate that surface proteins change their amino acid sequence to evade human immunity and these pressures can be recapitulated both in vivo and in vitro \citep{lee_mapping_2019,xue_within-host_nodate}.
A recent preprint proposed that influenza virus evolution is primarily limited by an asynchrony between population level selection and generation of new variants within infected hosts \citep{morris_asynchrony_2020}. 
Along these lines, it is possible that the A/A/H3N2 population readily responds once population level selection is high enough by giving rise to essentially equivalent variants.
In addition, every successful variant will produce population immunity limiting its future spread. 
These considerations might explain the disconnect between models of rapid adaptation and the frequency dynamics observed in influenza virus populations.

@@

\subsection*{Assigning a fitness to trajectories} % (fold)
\label{sub:assigning_a_fitness_to_trajectories}

% subsection assigning_a_fitness_to_trajectories (end)

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./SM_figures/pfix_v_freq_strat_h3n2_cutoff159S.png}
	\caption{Equivalent to figure \ref{fig:pfix_stratified} of the main text, but with strains carrying the HA1 159S mutation removed.}
	\label{fig:cutoff_159S_p2}
\end{figure*}

% \newpage
% \section{A/H1N1 influenza}
